---
title: "Untitled"
author: "Emmanouil Tranos"
date: "2022-09-09"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Street network analysis has been central to network science since its
nascence: its mathematical foundation, graph theory, was born in the
18th century when Leonhard Euler presented his famous Seven Bridges
of Königsberg problem.

Graphs can
also have parallel (i.e., multiple) edges between the same two nodes.

network may be thought of as a
real-world graph

A complex network is one with a
nontrivial topology (the configuration and structure of its nodes and
edges) – that is, the topology is neither fully regular nor fully random.
Most large real-world networks are complex (Newman, 2010). Of
particular interest to this study are complex spatial networks – that is,
complex networks with nodes and/or edges embedded in space
(O'Sullivan, 2014). A street network is an example of a complex spatial
networkwith both nodes and edges embedded in space, as are railways,
power grids, and water and sewage networks (Barthélemy, 2011).

street networks are approximately planar

Average street length

Node density

The eccentricity of a node is the maximum of the shortest-path
weighted distances between it and each other node in the network

The diameter of a network is the maximum eccentricity of any
node in the network and the radius of a network is theminimumeccentricity
of any node in the network. The center of a network is the node or
set of nodes whose eccentricity equals the radius and the periphery of a
network is the node or set of nodeswhose eccentricity equals the diameter

node degree

clustering coefficient

The maximum betweenness
centrality in a network specifies the proportion of shortest paths that
pass through themost important node/edge. This is an indicator of resilience:
networkswith a highmaximumbetweenness centrality aremore
prone to failure or inefficiency should this single choke point fail.

Closeness centrality represents, for each node, the reciprocal of the
sum of the distance from this node to all others in the network: that
is, nodes rank as more central if they are on average closer to all other
nodes. 

Finally, PageRank – the algorithm Google uses to rank web
pages – is a variant of network centrality, namely eigenvector centrality
(Brin & Page, 1998). PageRank ranks nodes based on the structure of incoming
links and the rank of the source node, andmay also be applied to
street networks (Agryzkov, Oliver, Tortosa, & Vicent, 2012; Chin &Wen,
2015)

OpenStreetMap – a collaborative mapping project that provides a
free and publicly editable map of the world – has emerged in recent
years as a major player both for mapping and for acquiring spatial
data (Corcoran, Mooney, & Bertolotto, 2013; Jokar Arsanjani, Zipf,
Mooney, & Helbich, 2015). Inspired by Wikipedia's mass-collaboration
model, the project started in 2004 and has grown to over two million
users today. Its data quality is generally quite high (Barron, Neis, &
Zipf, 2014; Girres & Touya, 2010; Haklay, 2010) and although data
coverage varies worldwide, it is generally good when compared to corresponding
estimates from the CIA World Factbook (Maron, 2015)

A city’s development eras, design paradigms, underlying terrain, culture, and local
economic conditions influence the pattern, topology, and grain of its street networks
(Jackson 1985; Kostof 1991). These networks in turn structure the human
interactions and transportation processes that run along them, forming an
important pillar of city planners’ quest for spatial order (Rose-Redwood and
Bigon 2018). In particular, network orientation and geometry have played an outsized
role in urban planning since its earliest days (Smith 2007).
Measuring these network patterns can help researchers, planners, and community
members understand local histories of urban design, transportation planning, and
morphology; evaluate existing transportation system patterns and configurations; and
explore new infrastructure proposals and alternatives. It also furthers the science of cities
by providing a better understanding of urban patterns and how they correspond to
evolutionary mechanisms, planning, and design.


```{r}
library(sfnetworks)
library(sf)
library(tidygraph)
library(tidyverse)
library(igraph)
```

```{r}
p1 = st_point(c(7, 51))
p2 = st_point(c(7, 52))
p3 = st_point(c(8, 52))
p4 = st_point(c(8, 51.5))

l1 = st_sfc(st_linestring(c(p1, p2)))
l2 = st_sfc(st_linestring(c(p1, p4, p3)))
l3 = st_sfc(st_linestring(c(p3, p2)))

edges = st_as_sf(c(l1, l2, l3), crs = 4326)
nodes = st_as_sf(c(st_sfc(p1), st_sfc(p2), st_sfc(p3)), crs = 4326)

edges$from = c(1, 1, 3)
edges$to = c(2, 3, 2)

net = sfnetwork(nodes, edges)
#> Checking if spatial network structure is valid...
#> Spatial network structure is valid
net
```

```{r}
nodes$name = c("city", "village", "farm")
edges$from = c("city", "city", "farm")
edges$to = c("village", "farm", "village")

edges

net = sfnetwork(nodes, edges, node_key = "name")
```

```{r}
st_geometry(edges) = NULL

other_net = sfnetwork(nodes, edges, edges_as_lines = TRUE)

plot(net, cex = 2, lwd = 2, main = "Original geometries")
plot(other_net, cex = 2, lwd = 2, main = "Straight lines")

```

```{r}

roxel

net = as_sfnetwork(roxel)
plot(net)

autoplot(net) + ggtitle("Road network of Münster Roxel")
```

```{r}
net = net %>%
  activate("nodes") %>%
  mutate(bc = centrality_betweenness())
#> Warning in betweenness(graph = graph, v = V(graph), directed = directed, :
#> 'nobigint' is deprecated since igraph 1.3 and will be removed in igraph 1.4

ggplot() +
  geom_sf(data = st_as_sf(net, "edges"), col = "grey50") +
  geom_sf(data = st_as_sf(net, "nodes"), aes(col = bc), size = 3) + #, size = bc
  ggtitle("Betweenness centrality in Münster Roxel")
```