---
title: "Network models in practice"
author: |
  | 
  | Emmanouil Tranos
  |
  | University of Bristol, Alan Turing Institute 
  | [e.tranos@bristol.ac.uk](mailto:e.tranos@bristol.ac.uk), [\@EmmanouilTranos](https://twitter.com/EmmanouilTranos),  [etranos.info](https://etranos.info/)
  | 
  | 
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
bibliography: "`r system('kpsewhich ../lectures/references.bib', intern = TRUE)`"
---

## Setup

Pay attention on the packages `sand` and `igraphdata`. The first one contains all 
the different network datasets used by [@kolaczyk2020statistical]. 
Then, the second one contain network data to be used together with the `igraph` package.
More information can be found [here](https://cran.r-project.org/web/packages/sand/sand.pdf) 
and [here](https://cran.r-project.org/web/packages/igraphdata/igraphdata.pdf). 
Feel free to explore these network data on your own time.

```{r setup}
library(tidyverse)
library(igraph)

library(sand)       
library(igraphdata)
```

## Aims of the practical

-   Simulate network models

-   Analyse their key attributes 

-   Compare them

`igraph` enables us to generate networks based on some of the some key network models
we discussed. I list below some of these functions:

-   `sample_gnp()`

-   `sample_smallworld()`

-   `sample_pa()`

Go to `igraph`'s [webpage](https://igraph.org/r/html/latest/) and familiarise yourselves
with the syntax of these functions.

```{block, type='alert alert-warning'}

**Task**: Create three networks using these three functions with, let's say, $n = 500$ nodes each.

Visualise and compare these networks.
```

Importantly, just before you generate a network with one of the above three functions
you will need to define a random seed for reproducibility reasons, e.g. `set.seed(55)`.

```{block, type='alert alert-warning'}

**Question**: *Without* defining a random seed, generate a network as per the above, 
run some type of quick analysis about this network and then repeat the code by generating 
a second network with the exact same characteristics. Run the same analysis for the
second network. Are the results the same?

Why do we need to define the random seed ?
```

Examples of modelled networks

```{r}

sample_gnp(n = 500, p = 0.02) %>% 
  plot(layout=layout_in_circle, vertex.label=NA)

sample_pa(n = 500, directed=FALSE) %>% 
  plot(layout=layout_in_circle, vertex.label=NA)

sample_smallworld(dim = 1, size = 500, nei = 5, p = 0.05) %>% 
  plot(layout=layout_in_circle, vertex.label=NA)

make_lattice(dim =1, length = 100, nei = 5) %>% 
  plot(vertex.label=NA)

```







